---
title: "How to retrieve data from the INBOVEG database"
author: "Els De Bie, Hans Van Calster, Jo Loos"
date: 2020-07-03
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_data_inboveg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

# Introduction

The Flemish vegetation database, INBOVEG, is an application developed to provide a repository of relevés and makes the relevés available for future use. More information? Check out https://www.vlaanderen.be/inbo/en-gb/data-applications/inboveg/

INBOVEG supports different types of recordings: BioHab recordings (protocol of Natura 2000 monitoring) and the classic relevés. 
The classic relevés can stand alone, be an element of a collection or element of a chain where the linkage is used to give information about the relative position of recording within a series. 
Ample selection and export functions toward analysis tools are provided. It also provides standardized lists of species, habitats, life forms, scales etc. 
Original observations are preserved and a full history of subsequent identifications is saved.



# Aim

We make functions available to query data directly from the INBOVEG SQL-server database. 
This avoids writing your own queries or to copy/paste them from the access-frontend for INBOVEG.

We have provided functions to query

* survey (INBOVEG-projects)
* metadata of recordings (header info)
* recordings (vegetation relevés)
* classification (Natura2000 or local classification like BWK)
* qualifiers (management and site characteristics)
* relations between recordings (relation information on recordings for one or more surveys based on Parent (classic-chain/bucket) and Child (classic) relationships)
  
  
# Packages and connection

The main functions that we will use in this tutorial all start with `get_inboveg_*`. 
These functions are made available by loading the `inbodb` package.

```{r setup}
library(inbodb)
```

These functions will only work for people with access to the INBO network. As an INBO employee, you should make sure you have reading-rights for CYDONIA, otherwise place an ICT-call.

The following R-code can be used to establish a connection to INBOVEG by using 'connect_inbo_dbase' of the inbodb-package with the database 'Cydonia' on the inbo-sql07-prd server:

```{r connection_inbo}
con <- connect_inbo_dbase("D0010_00_Cydonia")
```

# Functionality

## Survey information
The function `get_inboveg_survey` queries the INBOVEG database for survey information (metadata about surveys) for one or more survey(s) by the name of the survey.

Three examples are given, this can be used as base to continue selecting the data you require.

Get information of a specific survey and collect data.

```{r examples_survey}
survey_info <- get_inboveg_survey(con, 
                              survey_name = "OudeLanden_1979", 
                              collect = TRUE)
```

```{r}
survey_info
#> # A tibble: 1 x 5
#>      Id Name        Description                                   Owner creator 
#>   <int> <chr>       <chr>                                         <chr> <chr>   
#> 1   172 OudeLanden~ Verlinden A, Leys G en Slembrouck J (1979) G~ NA    els_deb~
```

Get information of all surveys. 
This time we will not use `collect = TRUE`, which will return a [lazy query](https://docs.lucee.org/guides/cookbooks/lazy_queries.html):

```{r}
allsurveys <- get_inboveg_survey(con)
```

```{r}
class(allsurveys)
#> [1] "tbl_Microsoft SQL Server" "tbl_dbi"                 
#> [3] "tbl_sql"                  "tbl_lazy"                
#> [5] "tbl" 
```


If only a part of the survey name is known, you can make use of wildcards such as `%`.

```{r}
partsurveys <- get_inboveg_survey(con, 
                              survey_name = "%MILKLIM%",
                              collect = TRUE)
```


## Header information
The function `get_inboveg_header` queries the INBOVEG database for header information (metadata for a vegetation-relevé) for one or more survey by the name of the survey(s) and the recorder type (classic, classic-emmer, classic-ketting, biohab). 

Three examples are given, this can be used as base to continue selecting the data you require.

Get header information from a specific survey and a specific recording type and collect the data:

```{r examples_header}
header_info <- get_inboveg_header(con, 
                              survey_name = "OudeLanden_1979",
                              rec_type = "Classic",
                              collect = TRUE)
```

```{r}
head(header_info, 10)
#> # A tibble: 10 x 15
#>    RecordingGivid Name  UserReference Observer LocationCode Latitude Longitude  Area
#>    <chr>          <chr> <chr>         <chr>    <chr>           <dbl>     <dbl> <dbl>
#>  1 IV20160210121~ Oude~ 5             Alex Ve~ Ekeren            500       500    10
#>  2 IV20160210140~ Oude~ 1             Alex Ve~ Ekeren            500       500    25
#>  3 IV20160210142~ Oude~ 2             Alex Ve~ Ekeren            500       500    25
#>  4 IV20160210153~ Oude~ 4             Alex Ve~ Ekeren            500       500    12
#>  5 IV20160210155~ Oude~ 81            Alex Ve~ Ekeren            500       500     9
#>  6 IV20160210160~ Oude~ 20            Alex Ve~ Ekeren            500       500     9
#>  7 IV20160210161~ Oude~ 17            Alex Ve~ Ekeren            500       500    12
#>  8 IV20160210162~ Oude~ 19            Alex Ve~ Ekeren            500       500    12
#>  9 IV20160210163~ Oude~ 18            Alex Ve~ Ekeren            500       500    12
#> 10 IV20160210164~ Oude~ 23            Alex Ve~ Ekeren            500       500     6
#> # ... with 7 more variables: Length <int>, Width <int>, VagueDateType <chr>,
#> #   VagueDateBegin <chr>, VagueDateEnd <chr>, SurveyId <int>, RecTypeID <int>
```

Get header information from several specific surveys by using `multiple = TRUE`. 

```{r}
header_severalsurveys <- get_inboveg_header(con,
                survey_name = c("MILKLIM_Heischraal2012", "NICHE Vlaanderen"),
                multiple = TRUE)
```

Get header information of all surveys without collecting the data:

```{r}
all_header_info <- get_inboveg_header(con)
```


## Recording information
The function  `get_inboveg_recordings` queries the INBOVEG database for relevé information (which species were recorded in which plots and in which vegetation layers with which cover) for one or more surveys.

Several examples are given, this can be used as base to continue selecting the data you require.

First, we show how to get the relevés from one survey and collect the data:

```{r examples_recordings}
recording_heischraal2012 <- get_inboveg_recordings(con, 
                                      survey_name = "MILKLIM_Heischraal2012",
                                      collect = TRUE)
```

```{r}
recording_heischraal2012
#> # A tibble: 3,288 x 11
#>    Name  RecordingGivid UserReference LayerCode CoverCode OrignalName ScientificName
#>    <chr> <chr>          <chr>         <chr>     <chr>     <chr>       <chr>         
#>  1 MILK~ IV20120816113~ HS_1001       M         5         Rhytidiade~ Rhytidiadelph~
#>  2 MILK~ IV20120816113~ HS_1001       M         5         Pseudoscle~ Pseudosclerop~
#>  3 MILK~ IV20120816113~ HS_1001       K         90        Juncus acu~ Juncus acutif~
#>  4 MILK~ IV20120816113~ HS_1001       K         90        Nardus str~ Nardus strict~
#>  5 MILK~ IV20120816113~ HS_1001       K         90        Potentilla~ Potentilla er~
#>  6 MILK~ IV20120816113~ HS_1001       K         90        Anthoxanth~ Anthoxanthum ~
#>  7 MILK~ IV20120816113~ HS_1001       K         90        Molinia ca~ Molinia caeru~
#>  8 MILK~ IV20120816113~ HS_1001       K         90        Lysimachia~ Lysimachia vu~
#>  9 MILK~ IV20120816113~ HS_1001       K         90        Luzula mul~ Luzula multif~
#>  10 MILK~ IV20120816113~ HS_1001       K         90        Carex pilu~ Carex pilulif~
#>  # ... with 3,278 more rows, and 4 more variables: PhenologyCode <chr>,
#>  #   CoverageCode <chr>, PctValue <dbl>, RecordingScale <chr>
```

Here is how to get the relevés for a specific RecordingGivid:

```{r}
recording_specific_givid <- get_inboveg_recordings(
  con, 
  recording_givid = "IV2012080609161322",
  collect = TRUE)
```

Or for multiple RecordingGivid, make sure to set `multiple = TRUE`. 

```{r}
recording_specific_givid <- get_inboveg_recordings(
  con, 
  recording_givid = c("IV2012080609161322", "IV2012081611384756"),
  multiple = TRUE,
  collect = TRUE)
```

The above is very useful in tandem with `get_inboveg_header()` to first search for all relevés belonging to one or more surveys and further filter these records to only the relevés you require based on the relevé metadata (such as date, observer, ...).


Or based on a user reference:

```{r}
recording_specific_user_reference <- get_inboveg_recordings(
  con, 
  user_reference = "HS_1001",
  collect = TRUE)
```

Get all recordings from MILKLIM surveys (partial matching), without collecting the data:

```{r}
recording_milkim <- get_inboveg_recordings(con,
                              survey_name = "%MILKLIM%")
```

Get recordings from several specific surveys:

```{r}
recording_severalsurveys <- get_inboveg_recordings(con,
                survey_name = c("MILKLIM_Heischraal2012", "NICHE Vlaanderen"),
                multiple = TRUE,
                collect = TRUE)
```

Get all relevés of all surveys,  without collecting the data:

```{r}
allrecordings <- get_inboveg_recordings(con)
```


## Classification information
The function `get_inboveg_classification` queries the INBOVEG database for information on the field classification (N2000 or BWK-code) of the relevé for one or more survey(s) by the name of the survey. 

Two  examples are given, this can be used as base to continue selecting the data you require.


Get a specific classification from a survey and collect the data:

```{r examples_classification}
classif_info <- get_inboveg_classification(con, 
                              survey_name = "MILKLIM_Heischraal2012",
                              classif = "4010",
                              collect = TRUE)
```

```{r}
classif_info
#> # A tibble: 1 x 9
#>   RecordingGivid Name  Classif ActionGroup ListName LocalClassifica~ Habitattype
#>   <chr>          <chr> <chr>   <chr>       <chr>    <chr>            <chr>      
#> 1 IV20130318144~ MILK~ 4010    N2k         Habitat~ NA               Noord-Atla~
#> # ... with 2 more variables: Cover <chr>, PctValue <dbl>
```


Get all surveys, all classifications without collecting the data:

```{r}
all_codes <- get_inboveg_classification(con)
```


## Qualifiers information
The function `get_inboveg_qualifiers`queries the INBOVEG database for qualifier information on recordings for one or more surveys. 
These qualifiers give information on management (management qualifier 'MQ') or location description (site qualifier'SQ').

Get the qualifiers from one survey:

```{r examples_qualifiers}
qualifiers_heischraal2012 <- get_inboveg_qualifiers(con,
                                  survey_name = "MILKLIM_Heischraal2012")
```

```{r}
head(qualifiers_heischraal2012)
#>                     Name     RecordingGivid UserReference
#>  1 MILKLIM_Heischraal2012 IV2012081615083167        HS_008
#>  2 MILKLIM_Heischraal2012 IV2012081613133274        HS_009
#>  3 MILKLIM_Heischraal2012 IV2013041614525228        HS_035
#>  4 MILKLIM_Heischraal2012 IV2012081609450300        HS_036
#>  5 MILKLIM_Heischraal2012 IV2012081610204607        HS_037
#>  6 MILKLIM_Heischraal2012 IV2012081610393743        HS_044
#>                                            Observer QualifierType Q1Code
#>  1                   Cécile Herr en Robin Guelinckx            MQ      A
#>  2                   Cécile Herr en Robin Guelinckx            MQ      A
#>  3 Cécile Herr, Patrik Oosterlynck, Robin Guelinckx            MQ      A
#>  4                   Cécile Herr en Robin Guelinckx            MQ      A
#>  5 Cécile Herr, Robin Guelinckx, Patrik Oosterlynck            MQ      A
#>  6                   Cécile Herr en Robin Guelinckx            MQ      A
#>    Q1Description Q2Code Q2Description   Q3Code Q3Description Elucidation NotSure
#>  1          <NA>  PBuis    Peilbuizen GUPP042A          <NA>                   0
#>  2          <NA>  PBuis    Peilbuizen GUPP043B          <NA>                   0
#>  3        Active  PBuis    Peilbuizen WALP161X          <NA>                   0
#>  4          <NA>  PBuis    Peilbuizen WALP157X          <NA>                   0
#>  5          <NA>  PBuis    Peilbuizen WALP117X          <NA>                   0
#>  6          <NA>  PBuis    Peilbuizen WALP162X          <NA>                   0
#>    ParentID  QualifierResource
#>  1       NA               <NA>
#>  2       NA               <NA>
#>  3       NA RS2012060811060080
#>  4       NA               <NA>
#>  5       NA               <NA>
#>  6       NA               <NA>
```


Get all site qualifiers (SQ) from MILKLIM surveys (partial matching):

```{r}
qualifiers_milkim <- get_inboveg_qualifiers(con,
                                        survey_name = "%MILKLIM%",
                                        qualifier_type = "SQ")
```

Get qualifiers from several specific surveys with `multiple = TRUE`:

```{r}
qualifiers_severalsurveys <- get_inboveg_qualifiers(con, 
                survey_name = c("MILKLIM_Heischraal2012", "NICHE Vlaanderen"),
                multiple = TRUE)
```


Get all qualifiers of all surveys:

```{r}
allqualifiers <- get_inboveg_qualifiers(con)
```


## Parent-child relationships between recordings

The function `get_inboveg_relation` queries the INBOVEG database for relational information on recordings for one or more surveys based on parent (classic-chain/bucket) and Child (classic) relationships.

Get all parent-child-relations from N2000meetnet surveys (partial matching):

```{r}
relations_N2000meetnet <- get_inboveg_relation(con, survey_name = "%N2000meetnet%")
```

```{r}
relations_N2000meetnet
#> # Source:   SQL [?? x 7]
#> # Database: Microsoft SQL Server
#> #   13.00.5598[INBO\hans_vancalster@INBO-SQL07-PRD\LIVE/D0010_00_Cydonia]
#>    Name   RecordingId Child_GIVID Child_UserRef ParentId Parent_GIVID Parent_UserRef
#>    <chr>        <int> <chr>       <chr>            <int> <chr>        <chr>         
#>  1 N2000~       44283 IV20190218~ 13594417_2170    44282 IV201902180~ 13594417_2170 
#>  2 N2000~       44285 IV20190218~ 10334001_2190    44284 IV201902181~ 10334001_2190 
#>  3 N2000~       44287 IV20190218~ 10186545_213~    44286 IV201902181~ 10186545_2130~
#>  4 N2000~       44290 IV20190218~ 10186545_2170    44289 IV201902181~ 10186545_2170 
#>  5 N2000~       44292 IV20190218~ 13561649_2170    44291 IV201902181~ 13561649_2170 
#>  6 N2000~       44294 IV20190218~ 11595569_2170    44293 IV201902181~ 11595569_2170 
#>  7 N2000~       44296 IV20190218~ 1240881_2170     44295 IV201902181~ 1240881_2170  
#>  8 N2000~       44298 IV20190218~ 13823793_2170    44297 IV201902181~ 13823793_2170 
#>  9 N2000~       44300 IV20190218~ 12644145_2170    44299 IV201902181~ 12644145_2170 
#> 10 N2000~       44302 IV20190218~ 10260273_213~    44301 IV201902181~ 10260273_2130~
#> # ... with more rows
```


Get the parent-child-relations from several specific surveys:

```{r}
relations_severalsurveys <- get_inboveg_relation(
  con, 
  survey_name =  c("DeBlankaart-1985-Beheer", "N2000meetnet_Grasland"), 
  multiple = TRUE)
```



## More complex queries
These functions give most of the relevant basic information that is available in INBOVEG.
If more detailed information is needed `dplyr` is the magic word. 
In the future additional functions can be added to this package to help the inboveg-users.




# Closing the connection

Close the connection when done

```{r closing_conn}
dbDisconnect(con)
rm(con)
```


