---
title: "How to retrieve data from the meetnetten database"
author: "Toon Westra, Frederic Piesschaert"
date: 2024-06-25
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to retrieve data from the meetnetten database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
tryCatch(
  con <- inbodb::connect_inbo_dbase("S0008_00_Meetnetten"),
  error = function(e) e,
  finally = database_access <- exists("con")
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = database_access,
  cache = FALSE
)
if (!database_access) {
  load("get_data_meetnetten_results.Rda")
}
```

# Introduction

The Flemish species monitoring programme [meetnetten.be](www.meetnetten.be) 
consists of a set of monitoring schemes for priority species in Flanders, the 
northern part of Belgium. 
Each monitoring scheme consists of a number of monitoring sites at which target 
species are counted using a standardized protocol. 
The count data is stored in the meetnetten database which is maintained by the
Nature and Forest Research Institute (INBO).
We present some functions to easily access the data from the meetnetten 
database.

# Aim

We make functions available to query data directly from the meetnetten database. 
This avoids writing your own queries.

We have provided functions to query

+ monitoring schemes included in the species monitoring programme
+ locations and sublocations (for example different sections of a transect) of 
the monitoring schemes
+ visits (count events)
+ observations
  
# Packages and connection

The main functions that we will use in this tutorial all start with 
`get_meetnetten_*`. 
These functions are made available by loading the `inbodb` package.

```{r setup, message = FALSE}
library(inbodb)
library(dplyr)
library(sf)
library(ggplot2)

```

These functions will only work for people with access to the INBO network. 
As an INBO employee, you should make sure you have read rights for the INBO 
Datawarehouse server, otherwise place an ICT-call.

The following R-code can be used to establish a connection to the 
meetnetten database:

```{r connection_inbo}
con <- connect_inbo_dbase("S0008_00_Meetnetten")
```

# Functionality

## Information on monitoring schemes

The function `get_meetnetten_schemes` gives an overview per species group of all
monitoring schemes included in meetnetten and the protocols that are applied in 
the monitoring schemes.

In most cases, a monitoring scheme is dedicated to one target species and the
monitoring scheme is named after the target species.
In some cases, a monitoring scheme has several target species, such as the 
Algemene Broedvogelmonitoring Vlaanderen. 


```{r get_scheme}
scheme_info <- get_meetnetten_schemes(con)
```

```{r eval=TRUE}
scheme_info
```

## Locations and sublocation of monitoring schemes

Each monitoring scheme consists of a fixed set of locations where the target
species is/are counted on a regular basis.
Some locations are subdivided in sublocations.
This is, for example, the case for most butterfly monitoring schemes, where the 
sublocations represent the 50 meter sections of a transect.

The function `get_meetnetten_locations` returns the locations and the
sublocations for one or several monitoring scheme.
It returns a list of two `sf` objects: `main_locations` and `sublocations`.

In the following example we select the locations and sublocations of the
Argusvlinder (*Lasiommata megera*) monitoring scheme.

```{r example_locations}

locations <- get_meetnetten_locations(con,
                                      scheme_name = "Argusvlinder")

main_locations <- locations$main_locations

sublocations <- locations$sublocations

```

Let's have a look at the `main_locations`.

```{r, eval = TRUE}
main_locations %>%
  head(5)
```
Two variables need some further explanations:

+ `is_sample`. For more common species, a random sample is drawn from all
locations in Flanders where the target species occurs.
For location that are selected `is_sample` = `TRUE`.
Locations that are not selected, indicated by `is_sample` = `FALSE`, can be
counted optionally.

+ `is_active`. When `is_active` = `FALSE`, the location is not counted any more.
This is the case when the location appears to be inaccessible or when the target
species does not occur at the location any more.
However, most often these inactive locations where counted in previous years.
So the the database also contains observations for (currently) inactive
locations. 

Next, let's have a look at the sublocations of the location Arenbergpolder.
The locations consists of 14 sections.
Note that we only show the active (`is_active` = `TURE`) sublocations. 

```{r, eval = TRUE}
sublocations_show <- sublocations %>%
  filter(location == "Arenbergpolder") %>%
  filter(is_active)

sublocations_show
```

Below we show a basic map of the locations and sublocations.
The location is the wider area where the target species occurs.
The sublocations indicate where the target species is counted.

```{r, fig.width = 7, eval = TRUE}

main_location_show <- main_locations %>%
  filter(location == "Arenbergpolder") 

ggplot(sublocations_show) +
  geom_sf(data = main_location_show) +
  geom_sf(aes(colour = sublocation))
```

We can also select and plot all locations from the dragonfly (in Dutch
*libellen*) monitoring schemes. 

```{r}
# get locations for a specific species_group
locations_dragonflies <- get_meetnetten_locations(con,
                                                  species_group = "libellen")

```

```{r, fig.width = 7, eval = TRUE}

locations_dragonflies$main_locations %>%
  ggplot() +
  geom_sf(aes(fill = scheme, colour = scheme), alpha = 0.5) 
```

## Visits

The function `get_meetnetten_visits` returns all visits (count events) for 
one or more monitoring schemes.

Below we select all visits for the location Arenbergpolder of the Argusvlinder
monitoring scheme. 
We see that in most years the location is counted 6 times per year as demanded
by the [monitoring protocol](https://purews.inbo.be/ws/portalfiles/portal/17657883/MaesEtal_2019_MonitoringsprotocolDagvlindersVersie2.pdf).

```{r example_visits}

visits <- get_meetnetten_visits(con,
                                scheme_name = "Argusvlinder",
                                collect = TRUE)

```

```{r, eval = TRUE}
visits %>%
  filter(location == "Arenbergpolder") %>%
  select(location, visit_id, start_date, 
         start_time, visit_status)
```

Use `?get_meetnetten_visits` to get more information on the variables that are
returned by the function.

## Observations

The function `get_meetnetten_observations` provides all observations for one 
or more monitoring scheme.

Let's select all observations from one visit at the Arenbergpolder location
of the Argusvlinder monitoring scheme, by applying a filter based on the
variable `visit_id`.

```{r example_observations}

observations <- get_meetnetten_observations(con,
                                            scheme_name = "Argusvlinder",
                                            collect = FALSE) %>%
  filter(visit_id == 1080) %>%
  collect()

```

First we take a look at the observation of the target species.

```{r, eval = TRUE}

observations %>%
  filter(target_species) %>%
  select(location, sublocation, start_date, 
         scientific_name, count)
```
Optionally observers can also record other species that can be counted using the
same protocol.

The code below provides the total number of individuals per species,
including the non target species, that were counted in one visit at the
Arenberpolder location.

```{r, eval = TRUE}

observations %>%
  group_by(start_date, target_species, name_nl, scientific_name) %>%
  summarise(count_total = sum(count)) %>%
  ungroup()
```

Use `?get_meetnetten_observations` to get more information on the variables that
are returned by the function. 

An important note: before you start analysing the data from a certain monitoring
scheme, it is important to know how the monitoring schemes were designed and how
the data is organised.
So don't forget to check the monitoring protocol (available on the 
[INBO website](https://www.vlaanderen.be/inbo/zoeken/?q=monitoringsprotocol)).

## Closing the connection

Close the connection when done

```{r closing_conn}
dbDisconnect(con)
rm(con)
```

```{r echo=FALSE, include=FALSE}
rm(database_access)
save.image(file = "get_data_meetnetten_results.Rda")
```

# Data policy

Since many of the target species are sensitive species, we do not publicly make 
available the exact location of the observations.
Location information is generalized to 1, 5 or 10 km Universal Transverse 
Mercator (UTM) grid cells, when publishing the data on [GBIF](www.gbif.org).
The generalisation rules are the same as in 
[waarnemingen.be](www.waarnemingen.be).

Please respect this policy when you publish results based on the meetnetten
data.
When in doubt, contact [info@meetnetten.be](info@meetnetten.be).
